### 2015-04-26 papers/histogram/thesis/thesis.tex
updates to histogram thesis

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Several
generic Monte Carlo ``histogram methods'' for collecting statistics on
low energy system states have been developed \fixme{by whom?}, but
little or no literature exists on their systematic comparison and
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We propose to implement these
methods in Monte Carlo simulations of the square-well fluid, determine
appropriate performance metrics for their comparison, and identify
those methods which are most efficient and effective at sampling
otherwise improbable system states.

@@

Monte Carlo simulations (Section \ref{sec:mc_sim}) are a standard
method for studying the thermodynamic properties of a system. In the
most straightforward implementation of Monte Carlo simulations,
however, highly localized distributions of spheres are extremely
unlikely to occur, making such simulations impractical for studying
the critical point of the square-well fluid. Several generic
computational methods, called histogram methods (Section
\ref{sec:histogram_methods}), exist for dealing with such a problem;
namely, that interesting regions of a system's state space cannot be
sampled via standard Monte Carlo simulations in a reasonable amount of
time. There is no consensus, however, on the relative efficacy of
these methods. Furthermore, some methods, such as the Wang-Landau
method\cite{wang_landau}, require tuning several free parameters that
lack both ``obvious'' values and systematic means for assigning
them. This work will thus attempt to quantify and compare the
effectiveness of a few such methods.

@@

\begin{algorithm}[tb]
  \caption{Unbiased Monte Carlo fluid simulation}
  \label{alg:MC}
  \begin{alg}

  \item Construct an initial ``typical'' fluid configuration.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased Monte Carlo fluid
simulations. Such an algorithm is ``unbiased'' in the sense that it
collects statistics (i.e. data) on all valid system configurations
with equal probability. Statistics whose collection time scales as
$\mathcal O\p{1}$ with system size, meaning that increasing the number
of spheres $N$ in the simulation does not affect collection time, can
be collected after every move (defined in the algorithm). Statistics
whose collection time scales as $\mathcal O\p{N}$, meaning that
doubling $N$ doubles collection time, can be collected after each
iteration. In general, collection with $\mathcal O\sp{\chi\p{N}}$ time
scaling should not occur more often than once every $\chi\p{N}$ moves
(where $\chi\p{N}$ may be an arbitrary function of only $N$ void of
constant prefactors, e.g. $N\log N$, or $2^N$). These collection rules
ensure that scaling up simulations does not cause them to
asymptotically spend all of their time collecting statistics, and no
time actually simulating the fluid. Collected statistics are used to
find thermodynamic properties of the simulated fluid.

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
unbiased Monte Carlo) to sufficiently sample $R$, that is, accumulate
a statistically significant histogram $H\p{X\in R}$, in any reasonable
amount of time.

@@

Histogram methods provide a means to address unbiased Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:mc_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

  \item Construct an initial ``typical'' fluid configuration.

@@

@@

\subsection{The flat histogram (multi-canonical) method}
\label{sec:flat_historam}

@@

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. A naive way to implement the flat histogram method without
knowing a priori the density of states $D\p{E}$ involves first running
a simulation without weights (or with constant weights $w\p{E}=w_0$)
while collecting a histogram $H\p{E}$ of the observed energies
$E$. After some time (loosely defined), the histogram $H\p{E}$ may be
taken as an approximation of the density of states $\tilde D\p{E}$
(see Section \ref{sec:histogram_methods}), and used via
(\ref{eq:flat_weights}) to generate a weight array for an actual
simulation.

The problem with this implementation is that histogram methods are
generally necessary when some range of energies $R$ is inaccessible
via unbiased Monte Carlo simulations. In such a case, the energy
histogram $H\p{E\in R}$ after simulating for any reasonable amount of
time will be statistically insignificant or null, invalidating the
approximation $H\p{E}\approx\tilde D\p{E}$. This method will therefore
only ``flatten'' the histogram at energies $E$ which have been
sufficiently sampled for $H\p{E}$ to be statistically significant. The
end goal in implementing histogram methods, however, is precisely to
sample those energies that cannot be sufficiently sampled without
employing clever schemes to bias Monte Carlo simulations. The flat
histogram method therefore does not directly help us study the
square-well fluid. Nonetheless, the this method has pedagogical value
by enabling discussion of an ``optimal'' weight array for generating a
flat energy histogram in simulations. All algorithms to construct
$w\p{E}$ with the aim of producing a flat energy histogram should
converge on the same weight array, given by (\ref{eq:flat_weights}).

@@

\fixme{Explain, provide algorithm. This method is our own.}

\begin{algorithm}[H]
  \caption{Simple flat initialization}
  \label{alg:simple_flat}
\end{algorithm}

@@

\fixme{Explain the motivation behind the Wang-Landau method, and
  provide the algorithm. Talk about the free parameters and roughly
  what they control.}

\begin{algorithm}[H]
  \caption{Wang-Landau calculation of weights}
  \label{alg:wang_landau}
\end{algorithm}

% \subsubsection{Modifications}
% \label{sec:wang_landau_mod}

% \fixme{Talk about our modifications to Wang-Landau made to adapt it
% to the square-well fluid, as well as (if I have time) other research
% groups' modifications.}

@@

\begin{algorithm}[H]
  \caption{Finding weights by optimizing energy walker currents}
  \label{alg:optimized_ensemble}
\end{algorithm}
uu
\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

@@

\begin{algorithm}[H]
  \caption{Transition matrix Monte Carlo simulation}
  \label{alg:tmmc}
\end{algorithm}

@@

@@

% \subsection{End conditions}
% \label{sec:end_conditions}

>>>>

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased (i.e. Metropolis) Monte Carlo
simulations are incapable of sampling the low energy fluid states
which dominate the partition function at low (i.e. liquid-state)
temperatures. Several generic Monte Carlo ``histogram methods'' for
collecting statistics on low energy system states have been developed
\fixme{by whom?}, but little or no literature exists on their
systematic comparison and their application to the square-well
fluid. The square-well fluid in particular introduces application
challenges not manifest in traditional models for testing and
benchmarking such numerical techniques (e.g. the Ising model). We
propose to implement these methods in Monte Carlo simulations of the
square-well fluid, determine appropriate performance metrics for their
comparison, and identify those methods which are most efficient and
effective at sampling otherwise improbable system states.

@@

Monte Carlo simulations are a standard means for studying the
thermodynamic properties of a system. In the most straightforward
implementation of Monte Carlo fluid simulations (Section
\ref{sec:mc_sim}), however, highly localized distributions of fluid
elements are extremely unlikely to occur, making such simulations
impractical for studying the critical point of the square-well
fluid. Several generic computational methods, called histogram methods
(Section \ref{sec:histogram_methods}), exist for dealing with such a
problem; namely, that interesting regions of a system's state space
cannot be sampled via standard Monte Carlo methods in a reasonable
amount of time. There is no consensus, however, on the relative
efficacy of these methods. Furthermore, some methods, such as the
Wang-Landau method\cite{wang_landau}, require tuning several free
parameters that lack both ``obvious'' values and systematic means for
assigning them. This work will thus attempt to quantify and compare
the effectiveness of a few such methods.

@@

\begin{algorithm}[tb]
  \caption{Metropolis Monte Carlo fluid simulation}
  \label{alg:MC}
  \begin{alg}

  \item Construct an initial ``typical'' (i.e. non-ordered) fluid
    configuration. \fixme{explain how}

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased, Metropolis Monte
Carlo fluid simulations. Such an algorithm is ``unbiased'' in the
sense that it collects statistics (i.e. data) on all valid system
configurations with equal probability. Statistics whose collection
time scales as $\mathcal O\p{1}$ with system size, meaning that
increasing the number of spheres $N$ in the simulation does not affect
collection time, can be collected after every move (defined in the
algorithm). Statistics whose collection time scales as $\mathcal
O\p{N}$, meaning that doubling $N$ doubles collection time, can be
collected after each iteration. In general, collection with $\mathcal
O\sp{\chi\p{N}}$ time scaling should not occur more often than once
every $\chi\p{N}$ moves (where $\chi\p{N}$ may be an arbitrary
function of only $N$, void of constant prefactors, e.g. $N\log N$, or
$2^N$). These collection rules ensure that scaling up simulations does
not cause them to asymptotically spend all of their time collecting
statistics, and no time actually simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

Due to the fact that Metropolis Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
Metropolis Monte Carlo) to sufficiently sample $R$, that is,
accumulate a statistically significant histogram $H\p{X\in R}$, in any
reasonable amount of time.

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:mc_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

  \item Construct an initial typical fluid configuration.

@@

\fixme{mention that simulating with flat weights $w\p{E}=w_0$ is the
  same as running a Metropolis MC simulation.}

@@

\fixme{Make a note on storing the logarithm of weights and
  periodically rescaling them (resetting the zero on $\ln w$) to
  reduce numerical error.}

\subsection{The flat histogram (multi-canonical) method}
\label{sec:flat_histogram}

@@

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. Nonetheless, the this method has pedagogical value by
enabling discussion of an ``optimal'' weight array for generating a
flat energy histogram in simulations. All algorithms to construct
$w\p{E}$ with the aim of producing a flat energy histogram should
converge on the same weight array, given by (\ref{eq:flat_weights}).

\begin{algorithm}[b]
  \caption{A naive flat histogram method}
  \label{alg:flat_histogram}
  \begin{alg}

  \item Construct an initial typical fluid configuration.

  \item Simulate (without weights) for some time $\tau$, which may be
    real time, computer time, or a number of simulation iterations,
    collecting a histogram $H\p{E}$ of energy observations after every
    move.

  \item Set weights $w\p{E}=1/H\p{E}$. At energies with $H\p{E}=0$,
    set $w\p{E}=1$.

Algorithm \ref{alg:flat_histogram} provides a naive way to implement
the flat histogram method without knowing a priori the density of
states $D\p{E}$. This algorithm involves first running a simulation
without weights (or with constant weights $w\p{E}=w_0$) while
collecting a histogram $H\p{E}$ of the observed energies $E$. After
some time (loosely defined), the histogram $H\p{E}$ may be taken as an
approximation of the density of states $\tilde D\p{E}$ (see Section
\ref{sec:histogram_methods}), and used via (\ref{eq:flat_weights}) to
generate a weight array for an actual simulation. Algorithm
\ref{alg:flat_histogram} has one free parameter, which determines the
amount of time for which to collect an energy histogram $H\p{E}$.

The problem with this implementation of the flat histogram method is
that histogram methods are generally necessary when some range of
energies $R$ is inaccessible via Metropolis Monte Carlo
simulations. In such a case, the energy histogram $H\p{E\in R}$ after
simulating for any reasonable amount of time will be statistically
insignificant or null, invalidating the approximation
$H\p{E}\approx\tilde D\p{E}$. This method will therefore only
``flatten'' the histogram at energies $E$ which have been sufficiently
sampled for $H\p{E}$ to be statistically significant. The end goal in
implementing histogram methods, however, is precisely to sample those
energies that cannot be sufficiently sampled without employing clever
schemes to bias Monte Carlo simulations. The flat histogram method
therefore does not directly help us study the square-well fluid, but
it does provide the basis for the method in Section
\ref{sec:simple_flat}.

@@

The ``simple flat'' method, developed by ourselves, extends the
implementation of the flat histogram method discussed in Section
\ref{sec:flat_histogram} to an algorithm intended to be the simplest
viable broad energy histogram method. While we do not expect this
method to outperform other, more sophisticated methods, the simple
flat method should provide a lower bar and a standard of comparison
for the performance of other methods. Any nontrivial histogram method
should, at a minimum, consistently outperform the simple flat method
in order to be considered for any applications.

The idea behind the simple flat method is such: after simulating for
some time, the energy histogram $H\p{E}$ will have a peak, or some
other uneven distribution of observations, in a region of energies $R$
for which $H\p{E\in R}$ statistically significant. One can thus use
the existing histogram to construct weights which should flatten
$H\p{R}$. Subsequent simulations should then spend less time at
energies with the highest state densities, and more time at energies
with relatively low state densities. The energy histogram from such
simulations should therefore be statistically significant in a larger
range of energies, improving the available estimate for a density of
states with which to compute flat histogram weights. One could then
use the improved estimate to construct new weights, and repeat this
simulation and updating process until simulations satisfy some end
condition (in fact, all histograms will require end conditions, which
is discussed in Section \ref{sec:end_conditions}).

\begin{algorithm}[!b]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

  \item Construct an initial typical fluid configuration.

  \item Set $k=0$, set initial weights $w_0\p{E}=1$ for all $E$, and
    choose an initial number of iterations $n_0$ for which to
    simulate.

  \item Simulate for $n_k$ iterations with weights $w_k\p{E}$,
    collecting a histogram $H_k\p{E}$ of energy observations after
    every move.
    \label{alg:simple_flat_sim}

  \item If some predetermined initialization end condition has
    \emph{not} been met, then:
    \begin{enumerate}
    \item increment $k\leftarrow k+1$,
    \item set $n_k=u\p{n_{k-1}}$ with a predetermined function
      $u:u\p{n}\ge n$,
      \label{alg:simple_flat_f}
    \item update weights as $w_k\p{E}=w_{k-1}\p{E}/H_{k-1}\p{E}$ for
      all $E$ at which $H_{k-1}\p{E}\ne0$,
      \label{alg:simple_flat_weight_update}
    \item return to step \ref{alg:simple_flat_sim}.
    \end{enumerate}

Algorithm \ref{alg:simple_flat} provides an implementation of the
simple flat method. This algorithm has two free ``parameters'': the
initial number of iterations $n_0$ for which to simulate, and the
function $u\p{n}$ used in step \ref{alg:simple_flat_f}. With an
appropriate choice of the function $u\p{n}$, this algorithm should not
be particularly sensitive to the value of $n_0$.

The most interesting part of Algorithm \ref{alg:simple_flat} is step
\ref{alg:simple_flat_weight_update}, which updates the weights as
$w_k\p{E}=w_{k-1}\p{E}/H_{k-1}\p{E}$. The reasoning behind this step
is such: say that for two energies $E_0$ and $E_b$ we have that
$H\p{E_0}=\bk{H}_E$ and $H\p{E_b}=b\bk{H}_E$, where $\bk{H}_E$ is the
mean histogram value over all energies. In this case, we wish to
decrease the current simulation bias on $E_b$ by a factor of $b$ while
keeping the bias on $E_0$ the same. As the weights $w\p{E}$ are
proportional to the simulation bias on $E$, dividing the weights
$w\p{E}$ by the current histogram $H\p{E}$ achieves the desired
changes to the simulation biases. The factor $\bk{H}_E$ in this weight
update does not affect results, and is in practice immediately scaled
out of the weight array.

In our implementation of the simple flat method, we chose $u\p{n}=2n$
so that even if the choice of initial $n_0$ is too small, $n_k=2^kn_0$
should quickly approach the appropriate scale for the system at
hand. We also used $n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is
the number of energy levels in the given square-well fluid and
$T_{\t{min}}$ is a minimum temperature of interest for the current
simulation. The factor of $L$ appears because initialization time
should scale with the range of energies of the simulated system. If we
have twice the number of energies, we should explore energy space for
twice as long. To estimate $L$, we multiplied the maximum number of
spheres $M$ which fit within a radius of $\lambda\sigma$ of a single
sphere on a face-centered cubic lattice (i.e. the maximum number of
spheres with which a given sphere could interact) by the total number
of spheres $N$, and divided the result by two so as to not
double-count interactions between spheres (so $L=MN/2$). The
exponential $\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for
two adjacent energies, appears because lower minimum temperatures of
interest warrant sampling lower energies, which in turn requires
simulating for longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (see Section \ref{sec:min_energy}). In this paper,
we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

The Wang-Landau method is the first published histogram method we
discuss, and is the standard method\fixme{cite} to achieve broad
energy sampling in Monte Carlo simulations\cite{wang_landau,
  wang_landau_analysis, wang_landau_mod}.

\fixme{Explain the motivation behind the Wang-Landau method, and
  explain/provide the algorithm.}

\begin{algorithm}[tb]
  \caption{Wang-Landau initialization of weights}
  \label{alg:wang_landau}
  \begin{alg}

  \item Construct an initial typical fluid configuration

  \item Set $k=0$, choose some factor $f_0>0$, and initialize
    $w\p{E}=1$ for all $E$.

  \item Simulate for $n$ iterations. After each move, increment the
    histogram $H\p{E}$ of energy observations and update
    $w\p{E}\leftarrow w\p{E}/f_k$, where $E$ is the current energy of
    the system.
    \label{alg:wang_landau_sim}

  \item If the histogram $H\p{E}$ is sufficiently flat, i.e. if it
    satisfies some predetermined ``flatness'' condition $C\sp{H}$,
    then
    \begin{enumerate}[label=(\alph*)]
    \item increment $k\leftarrow k+1$,
    \item set $f_k=u\p{f_{k-1}}$ with a predetermined function
      $u:u\p{f}<f$,
    \item if $f_k>c$ for some predetermined $c$, return to step
      \ref{alg:wang_landau_sim}.
    \end{enumerate}

@@

\begin{algorithm}[H]
  \caption{Modifying weights via the optimized ensemble}
  \label{alg:optimized_ensemble}
\end{algorithm}

\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

@@

\begin{algorithm}[H]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
\end{algorithm}

@@

\fixme{Mention that in all initializations which simulate for some
  time with unchanging weights (i.e. simple flat), the weight array
  should be set to canonical weights below the minimum important
  energy, as soon as it is identified.}

\fixme{Mention that weights should be set to canonical weights below
  the minimum important energy after all initializations.}

@@

\subsection{End conditions}
\label{sec:end_conditions}
